---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# _demografr_: Simple and efficient Approximate Bayesian Computation in R

<!-- badges: start -->
<!-- badges: end -->

The goal of _demografr_ is to simplify and streamline [Approximate Bayesian Computation](https://en.wikipedia.org/wiki/Approximate_Bayesian_computation) (ABC) in population genetics and make it more reproducible. Additionally, _demografr_ aims to make ABC orders of magnitude faster and more efficient by leveraging the [tree sequences](https://tskit.dev/learn/) as an internal data structure and computation engine.

Unlike traditional ABC approaches, which generally involve custom-built pipelines and scripts for population genetic simulation and computation of summary statistics, _demografr_ makes it possible to perform simulation, data analysis, and ABC inference itself entirely in R within a single reproducible analysis script. By eliminating the need to write custom simulation code and scripting for integration of various population genetic tools for computing summary statistics, it lowers the barrier to entry for new users and facilitates reproducibility for all users regardless of their level of experience by eliminating many common sources of bugs.

### What are the issues with standard ABC?

A traditional ABC analysis in population genetics generally involves:

1. Writing a simulation script tailor-made for the demographic model of species in question using tools such as [ms](http://home.uchicago.edu/rhudson1/source/mksamples.html), [scrm](https://github.com/scrm/scrm/), [msprime](https://github.com/tskit-dev/msprime/), or [SLiM](https://github.com/MesserLab/SLiM/). For most realistic demographic models this presents a significant obstacle unless one is well versed in software development.
2. Developing a pipeline which will draw model parameters from a set of priors, and simulate output data in an appropriate format. Saving output files to disk can be very slow, especially for large genome-scale simulations across thousands of simulation replicates.
3. Computing summary statistics on the simulated data using appropriate software (often a combination of ADMIXTOOLS, PLINK, vcftools, or various custom scripts). This usually requires conversion of simulated outputs to an appropriate file format. In addition to the cost of disk-access, computing summary statistics can be quite slow depending on the statistic or program in question.
4. Feeding (and often re-formatting) the outputs of summary statistics in the right format to use with the appropriate ABC inference engine, such as the R package [_abc_](https://cran.r-project.org/package=abc).

Given the range of disparate tools required for the steps above, traditional ABC pipelines end up being quite complex, require a large amount of programming work and, because of the specifics of each demographic model, the whole procedure is usually re-invented from scratch for each study.

All in all, **an unnecessary amount of time spent on ABC analyses is dedicated to software development&mdash;programming, debugging, and data munging&mdash;work that would be better spent doing research**.

### How does _demografr_ help?

_demografr_ streamlines every step of a typical ABC pipeline by leveraging the [_slendr_](https://github.com/bodkan/slendr/) framework as a building block for simulation and data analysis, making it possible to perform an entire ABC analysis in R. Taking the steps above one by one, here is how _demografr_ makes life easier:

1. _slendr_'s intuitive, interactive [interface for definning population genetic models](https://www.slendr.net/articles/vignette-04-nonspatial-models.html) makes it easy to encode even complex demographic models with only bare minimum of R knowledge needed.
2. _demografr_ makes it possible to encode prior distributions of parameters using familiar R interface resembling standard probabilistic statements, and provides an automated function which simulates ABC replicates drawing parameters from priors in a trivial, one-step manner.
3. Because _slendr_'s simulation output is a [tree sequence](https://tskit.dev/learn/), most population genetic statistics can be computed directly on such tree sequences using R functions which are part of _slendr_'s statistical library. A tree sequence is never saved to disk and no conversion between file formats is required.
4. _demografr_ facilitates tight integration with the powerful R package [_abc_](https://cran.r-project.org/package=abc) by automatically feeding its simulation data to the _abc_ package for inference and analysis.

## Installation

You can install the development version of _demografr_ from [GitHub](https://github.com/) with:

``` r
devtools::install_github("bodkan/demografr")
```

Note that this requires an R package _devtools_, which you can obtain simply by running `install.packages("devtools")`.

### Note on stability

_demografr_ is very much in an early experimental stage at this point. Although ABC fitting of "standard" demographic models (i.e. estimating $N_e$, split times and gene-flow parameters for non-spatial models) already works very nicely, our long-term ambitions for the project are much higher. As such, please be aware that the interface might change significantly on a short notice to accomodate features for estimating parameters of more complex custom models such as spatial models etc.

## An example ABC analysis

```{r, echo=FALSE}
devtools::install(".", upgrade = "never")
```

```{r, results="hide"}
library(dplyr)
library(slendr)
library(demografr)
```

```{r setup, echo=FALSE}
SEED <- 42

p1 <- population("popA", time = 1, N = 1000)
p2 <- population("popB", time = 2000, N = 3000, parent = p1)
p3 <- population("popC", time = 4000, N = 10000, parent = p2)
p4 <- population("popD", time = 6000, N = 5000, parent = p3)

example_model <- compile_model(
  populations = list(p1, p2, p3, p4),
  generation_time = 1,
  simulation_length = 10000, serialize = FALSE
)

ts <- msprime(example_model, sequence_length = 100e6, recombination_rate = 1e-8, random_seed = SEED) %>% ts_mutate(1e-8, random_seed = SEED)

samples <- ts_samples(ts) %>% split(., .$pop) %>% lapply(`[[`, "name")

pi_df <- ts_diversity(ts, sample_sets = samples)  %>% mutate(stat = paste0("pi_", set))       %>% select(stat, value = diversity)

div_df <- ts_divergence(ts, sample_sets = samples)  %>% mutate(stat = sprintf("d_%s_%s", x, y))       %>% select(stat, value = divergence)
```

Imagine that we sequenced genomes of individuals from three populations "p1", "p2", and "p3".

Let's also assume that we know that the three populations are phylogenetically related in the following way but we don't know anything else (i.e., we have no idea about their $N_e$ or split times):

```{r ape_tree, echo=FALSE, fig.width=4, fig.height=3}
tree <- ape::read.tree(text="(popA,(popB,(popC,popD)));")
plot(tree)
```

After sequencing the genomes of individuals from these populations, we computed the nucleotide diversity in these populations as well as their pairwise genetic divergence, and observed the following values which we saved in two standard R data frames:

```{r}
pi_df
```

```{r}
div_df
```

For the purpose of the ABC analysis below, we will bind all statistics in to a standard R list, naming them appropriately:

```{r}
observed <- list(diversity = pi_df, divergence = div_df)
```

### Setting up a "scaffold" model

The first step in a _demografr_ ABC analysis is setting up a "scaffold" model for which we will estimate the posterior parameters of interest.

One way to do this is by building a normal _slendr_ model (for instance, a model such as [this](https://www.slendr.net/articles/vignette-04-nonspatial-models.html)). However, for simpler models like ours, it can be easier to input the scaffold as a standard phylogenetic tree (here we use the function `tree_model()` to input a tree in the [Newick format](https://en.wikipedia.org/wiki/Newick_format)).

Note that the parameter `time_span` indicates how much evolutionary time does of our model cover (this can be currently only specified in generations):

```{r}
model <- tree_model(tree = "(popA,(popB,(popC,popD)));", time_span = 10000)
```

Because the `model` object contains a standard _slendr_ demographic model, we can visualize it as such, to make sure everything is set up correctly:

```{r, slendr_model, fig.width=5, fig.height=3}
plot_model(model)
```

 Note that we don't pay attention to the split times or population sizes of this model because we will be fitting those parameters with the ABC procedure below&mdash;the model we just constructed is really just a _scaffold_ capturing some prior information we have about the phylogenetic relationship between populations. That said, we have the option to fix some aspects of our model by building and fine-tuning the model using standard _slendr_ features for defining models rather than importing the model as a phylogenetic tree (see [this](https://www.slendr.net/articles/vignette-04-nonspatial-models.html) _slendr_ vignette).

### Setting up priors

We are interested in estimating the $N_e$ of all populations and their split times. _demografr_ makes this very easy using a familiar symbolic formula syntax in R:

```{r}
priors <- list(
  Ne_popA ~ runif(1, 10000),
  Ne_popB ~ runif(1, 10000),
  Ne_popC ~ runif(1, 10000),
  Ne_popD ~ runif(1, 10000),

  Tsplit_popA_popB ~ runif(1, 3000),
  Tsplit_popB_popC ~ runif(3000, 6000),
  Tsplit_popC_popD ~ runif(6000, 9000)
)
```

In an ABC simulation step below, the formulas are used to draw the values of each parameter from specified distributions (in this case, all uniform distributions across a wide range of parameter values).

### Defining summary functions

Each run of a _demografr_ ABC simulation internally produces a tree sequence as an output. Because tree sequence represents an efficient, succint representation of the complete genealogical history of a set of samples, it is possible to compute population genetic statistics directly on the tree sequence without having to first save each simulation output to disk for computation in different software. Thanks to _slendr_'s library of [tree-sequence functions](https://www.slendr.net/reference/index.html#tree-sequence-statistics) serving as an R interface to the [_tskit_ module](https://tskit.dev/tskit/docs/stable/stats.html), you can specify summary statistics to be computed for ABC using plain and simple R code.

In our example, because we computed nucleotide diversity and pairwise divergence in the individuals sequenced from populations "p1", "p2", and "p3", we will define the following functions. Crucially, when run on a tree-sequence object, they will produce an output data frame in the format analogous to the empirical statistics shown in data frames `diversity` and `divergence` above:

```{r}
compute_diversity <- function(ts) {
  samples <- ts_samples(ts) %>% split(., .$pop) %>% lapply(`[[`, "name")
  ts_diversity(ts, sample_sets = samples) %>%
    mutate(stat = paste0("pi_", set)) %>%
    select(stat, value = diversity)
}

compute_divergence <- function(ts) {
  samples <- ts_samples(ts) %>% split(., .$pop) %>% lapply(`[[`, "name")
  ts_divergence(ts, sample_sets = samples) %>%
    mutate(stat = sprintf("d_%s_%s", x, y)) %>%
    select(stat, value = divergence)
}

functions <- list(diversity  = compute_diversity, divergence = compute_divergence)
```

Crucially, the outputs of these summary functions _must_ match the format of the observed summary statistics (i.e., the data frames produced must have the same format). This minor inconvenience during ABC setup saves us the headache of having to match values of statistics between observed and simulated data during ABC inference itself.

### ABC simulations

Having defined the scaffold model, a set of priors for our parameters of interest ($N_e$ and split times), as well as two summary statistic functions, we can plug all this information into the function `simulate_abc`.

Before we run a potentially computationally costly simulations, it is a good idea to validate the ABC components we have so far assembled using the function `validate_abc()`:

```{r}
validate_abc(model, priors, functions, observed)
```

Having verified that all model components are set up correctly, we can proceed to the ABC simulations themselves, using _demografr_'s function `simulate_abc()`:

```{r, eval=!file.exists("inst/extdata/data.rds")}
data <- simulate_abc(
  model, priors, functions, observed, iterations = 10000,
  sequence_length = 10e6, recombination_rate = 1e-8, mutation_rate = 1e-8
)

data
```

```{r, echo=FALSE}
if (file.exists("inst/extdata/data.rds")) {
  data <- readRDS("inst/extdata/data.rds")
} else {
  saveRDS(data, "inst/extdata/data.rds")
}
```

At this point we have generated summary statistics for simulations of models using parameters drawn from our priors. In the next step, we can finally do inference of our parameters.

### ABC inference

Having all the information about observed and simulated data bound in a single R object `abc_data`, we can finally perform the ABC inference. _demografr_ includes a convenient function `perform_abc()` which reformats the simulated and observed data in a format required by the R package [_abc_](https://cran.r-project.org/package=abc) and internally calls [the function `abc()`](https://cran.r-project.org/web/packages/abc/abc.pdf) of that package.

Note that `perform_abc` is just convenience wrapper around the `abc()` function in the package _abc_`, saving us a little work juggling the necessary matrices manually. As such, all parameters of the function `abc()` can be provided to `perform_abc()`, which will then pass them on appropriately.

```{r}
abc <- perform_abc(data, tolerance = 0.05, method = "neuralnet")
```

### Inspecting the posteriors

Now that we have the ABC output object ready, we can get a data frame with summary statistics of the posterior distributions of our parameters. For instance, we can easily read the maximum a posteriori probability (MAP) of the parameters in the row labelled "Weighted Mode:":

```{r}
extract_summary(abc)
```

Because a chart is always more informative than a table, we can easily get a visualization of our posteriors using the function `plot_posterior()`:

```{r, demografr_posterior_Ne}
plot_posterior(abc, type = "Ne")
```

```{r, demografr_posterior_Tsplit}
plot_posterior(abc, type = "Tsplit")
```

Because the internals of _demografr_ ABC objects are represented by standard objects created by the _abc_ package, we have many of the standard diagnostics functions of the _abc_ R package at our disposal:

```{r, abc_diagnostic_Ne}
plot(abc, param = "Ne_popC")
```

## The example as a single script

```{r, eval=FALSE}
model <- tree_model(tree = "(popA,(popB,(popC,popD)));", time_span = 10000)

priors <- list(
  Ne_popA ~ runif(1, 10000),
  Ne_popB ~ runif(1, 10000),
  Ne_popC ~ runif(1, 10000),
  Ne_popD ~ runif(1, 10000),

  Tsplit_popA_popB ~ runif(1, 3000),
  Tsplit_popB_popC ~ runif(3000, 6000),
  Tsplit_popC_popD ~ runif(6000, 9000)
)

compute_diversity <- function(ts) {
  samples <- ts_samples(ts) %>% split(., .$pop) %>% lapply(`[[`, "name")
  ts_diversity(ts, sample_sets = samples) %>%
    mutate(stat = paste0("pi_", set)) %>%
    select(stat, value = diversity)
}

compute_divergence <- function(ts) {
  samples <- ts_samples(ts) %>% split(., .$pop) %>% lapply(`[[`, "name")
  ts_divergence(ts, sample_sets = samples) %>%
    mutate(stat = sprintf("d_%s_%s", x, y)) %>%
    select(stat, value = divergence)
}

summary_funs <- list(diversity  = compute_diversity, divergence = compute_divergence)

```