---
title: "Genetic algorithm optimization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genetic algorithm optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
slendr_present <- slendr::check_dependencies(python = TRUE, slim = TRUE, quit = FALSE)

knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.align = "center",
  fig.width = 8,
  fig.height = 5,
  dpi = 80,
  eval = slendr_present
)

result_path <- here::here("inst/examples/ga_result.rds")
time_path <- here::here("inst/examples/ga_time.rds")

devtools::load_all()
```

```{r, echo=FALSE, message=FALSE}
library(demografr)
library(slendr)
init_env()

library(GA)

library(dplyr)
library(ggplot2)

SEED <- 42
set.seed(SEED)
```

Load empirical summary statistics computed from the observed data:

```{r}
observed_diversity <- read.table(system.file("examples/observed_diversity.tsv", package = "demografr"), header = TRUE)
observed_divergence <- read.table(system.file("examples/observed_divergence.tsv", package = "demografr"), header = TRUE)
observed_f4  <- read.table(system.file("examples/observed_f4.tsv", package = "demografr"), header = TRUE)

observed <- list(
  diversity  = observed_diversity,
  divergence = observed_divergence,
  f4         = observed_f4
)
```

Specify the scaffold model function (each function argument corresponds to one parameter to be inferred):

```{r}
model <- function(Ne_A, Ne_B, Ne_C, Ne_D, T_AB, T_BC, T_CD, gf_BC) {
  popA <- population("popA", time = 1,    N = Ne_A)
  popB <- population("popB", time = T_AB, N = Ne_B, parent = popA)
  popC <- population("popC", time = T_BC, N = Ne_C, parent = popB)
  popD <- population("popD", time = T_CD, N = Ne_D, parent = popC)

  gf <- gene_flow(from = popB, to = popC, start = 9000, end = 9301, rate = gf_BC)

  model <- compile_model(
    populations = list(popA, popB, popC, popD), gene_flow = gf,
    generation_time = 1, simulation_length = 10000,
    direction = "forward", serialize = FALSE
  )

  samples <- schedule_sampling(
    model, times = 10000,
    list(popA, 25), list(popB, 25), list(popC, 25), list(popD, 25),
    strict = TRUE
  )

  return(list(model, samples))
}
```

Specify tree-sequence summary statistics (one function per one observed statistic, as defined above):

```{r}
compute_diversity <- function(ts) {
  samples <- ts_names(ts, split = "pop")
  ts_diversity(ts, sample_sets = samples)
}
compute_divergence <- function(ts) {
  samples <- ts_names(ts, split = "pop")
  ts_divergence(ts, sample_sets = samples)
}
compute_f4 <- function(ts) {
  samples <- ts_names(ts, split = "pop")
  A <- samples["popA"]; B <- samples["popB"]
  C <- samples["popC"]; D <- samples["popD"]
  rbind(
    ts_f4(ts, A, B, C, D),
    ts_f4(ts, A, C, B, D),
    ts_f4(ts, A, D, B, C)
  )
}

functions <- list(
  diversity  = compute_diversity,
  divergence = compute_divergence,
  f4         = compute_f4
)
```


<!-- ```{r} -->
<!-- saveRDS(result, result_path) -->
<!-- saveRDS(time, time_path) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- result <- readRDS(result_path) -->
<!-- time <- readRDS(time_path) -->
<!-- ``` -->

```{r}
bounds <- list(
  Ne...  ~ between(1, 10000),
  T...   ~ between(1, 10000),
  gf_BC  ~ between(0, 1)
)

start <- Sys.time()
result <- run_ga(model, functions, observed, bounds, iterations = 500, error_fun = mean,
                 sequence_length = 50e6, recombination_rate = 1e-8, mutation_rate = 1e-8,
                 parallel = TRUE)
end <- Sys.time()
time <- end - start
print(time)
```

```{r}
saveRDS(result, result_path)
saveRDS(time, time_path)
```

```{r}
result <- readRDS(result_path)
```

```{r}
plot(result)
```

Checking the fitness of the best solution:

```{r}
colnames(result@solution) <- names(formals(model))
result@solution
```

```{r}
result@fitnessValue
```

```{r}
compute_fitness(
  result@solution, model, functions, observed,
  sequence_length = 25e6, recombination_rate = 1e-8, mutation_rate = 1e-8
)
```

Fitness value for the true parameter vector:

```{r}
params <- list(Ne_A = 2000, Ne_B = 800, Ne_C = 9000, Ne_D = 4000, T_AB = 2000, T_BC = 6000, T_CD = 8000, gf = 0.1)

compute_fitness(
  params, model, functions, observed,
  sequence_length = 25e6, recombination_rate = 1e-8, mutation_rate = 1e-8, statistics = TRUE
)
```

```{r}
params <- list(Ne_A = 20, Ne_B = 80, Ne_C = 90, Ne_D = 40, T_AB = 20, T_BC = 6000, T_CD = 8000, gf = 0.9)

compute_fitness(
  params, model, functions, observed,
  sequence_length = 25e6, recombination_rate = 1e-8, mutation_rate = 1e-8, statistics = TRUE
)
```

```{r}
params <- list(Ne_A = 2000, Ne_B = 800, Ne_C = 9000, Ne_D = 4000, T_AB = 2000, T_BC = 6000, T_CD = 8000, gf = 0.1)

stats <- parallel::mclapply(1:1000, function(i)
  compute_fitness(
    params, model, functions, observed, statistics = TRUE,
    sequence_length = 200e6, recombination_rate = 1e-8, mutation_rate = 1e-8
  )[[1]] %>% dplyr::mutate(rep = i), mc.cores = 100) %>% do.call(rbind, .)
```

```{r}
saveRDS(stats, "stats.rds")
```

```{r}
p_diversity <- stats %>% filter(stat == "diversity") %>%
  ggplot() +
  geom_density(aes(error, color = id))

p_divergence <- stats %>% filter(stat == "divergence") %>%
  ggplot() +
  geom_density(aes(error, color = id))

p_f4 <- stats %>% filter(stat == "f4") %>%
  ggplot() +
  geom_density(aes(error, color = id))
```

```{r}
cowplot::plot_grid(p_diversity, p_divergence, p_f4)
```


