---
title: "vignette-02-split_times"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette-01-overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)

devtools::install(".")
```

```{r setup}
library(dplyr)
library(slendr)
library(combinat)

library(demografr)
```

```{r}
# library(future)
# nodes <- c("r1", "r2", "r3", "r4")
# plan(list(tweak(cluster, workers = nodes, homogeneous = FALSE),
#           tweak(multisession, workers = 50)))

# plan(multicore, workers = 10)
```

Define a skeleton model to fit parameters for:

```{r}
p1 <- population("p1", time = 1,    N = 3000)
p2 <- population("p2", time = 2000, N = 800, parent = p1)
p3 <- population("p3", time = 6000, N = 5000, parent = p2)
p4 <- population("p4", time = 9000, N = 2000, parent = p3)

gf <- gene_flow(from = p4, to = p2, start = 9500, end = 9501, rate = 0.17)

model <- compile_model(
  populations = list(p1, p2, p3, p4), gene_flow = gf,
  generation_time = 1,
  simulation_length = 10000, serialize = FALSE
)

plot_model(model, proportions = TRUE)
```

Simulate true summary statistics from this model. Later, we will use those to fit parameters of the model, pretending that we forgot the truth:

```{r}
ts <- msprime(model, sequence_length = 100e6, recombination_rate = 1e-8) %>% ts_mutate(1e-8)

samples <- ts_samples(ts) %>% split(., .$pop) %>% lapply(pull, name)

observed_stats <- list(
  pi         = ts_diversity(ts, sample_sets = samples)  %>% mutate(stat = paste0("pi_", set))       %>% select(stat, value = diversity),
  divergence = ts_divergence(ts, sample_sets = samples) %>% mutate(stat = sprintf("d_%s_%s", x, y)) %>% select(stat, value = divergence),
  f4 = permn(samples) %>%
    lapply(function(pops) {
      ts_f4(ts, W = pops[1], X = pops[2], Y = pops[3], Z = pops[4]) 
    }) %>%
    do.call(rbind, .) %>%
    mutate(stat = paste("f4", W, X, Y, Z, sep = "_")) %>%
    select(stat, value = f4)
)
```

Define prior distributions for model parameters:

```{r}
priors <- list(
  Ne_p1 ~ runif(1, 10000),
  Ne_p2 ~ runif(1, 10000),
  Ne_p3 ~ runif(1, 10000),
  Ne_p4 ~ runif(1, 10000),

  Tsplit_p1_p2 ~ runif(1, 3000),
  Tsplit_p2_p3 ~ runif(3500, 7500),
  Tsplit_p3_p4 ~ runif(8000, 10000),

  gf_p4_p2 ~ runif(0, 1)
)

plot_prior(priors)
```

Define set of R functions operating on a tree sequence which will be used to compute summary statistics during the ABC inference:

```{r}
compute_pi <- function(ts) {
  samples <- ts_samples(ts) %>% split(., .$pop) %>% lapply(`[[`, "name")
  ts_diversity(ts, sample_sets = samples) %>%
    mutate(stat = paste0("pi_", set)) %>%
    select(stat, value = diversity)
}

compute_divergence <- function(ts) {
  samples <- ts_samples(ts) %>% split(., .$pop) %>% lapply(`[[`, "name")
  ts_divergence(ts, sample_sets = samples) %>%
    mutate(stat = sprintf("d_%s_%s", x, y)) %>%
    select(stat, value = divergence)
}

compute_f4 <- function(ts) {
  samples <- ts_samples(ts) %>% split(., .$pop) %>% lapply(pull, name)
  perms <- permn(samples)

  lapply(perms, function(pops) ts_f4(ts, W = pops[1], X = pops[2], Y = pops[3], Z = pops[4])) %>%
    do.call(rbind, .) %>%
    mutate(stat = paste("f4", W, X, Y, Z, sep = "_")) %>%
    select(stat, value = f4)
}

functions <- list(
  pi = compute_pi,
  divergence = compute_divergence,
  f4 = compute_f4
)
```

Generate simulation replicates for ABC analysis, drawing parameters of the model from the prior and computing summary statistics for each simulation:

```{r}
Sys.time()
data <- simulate_abc(
  model,                           # "scaffold" slendr model
  priors,                          # prior distributions
  summary_funs = functions,        # user-defined summary functions
  observed_stats = observed_stats, # observed popgen statistics
  iterations = 1000,               # number of ABC simulation replicates
  sequence_length = 10e6, recombination_rate = 1e-8, mutation_rate = 1e-8
)
Sys.time()

saveRDS(data, "data_gf.rds")
```

```{r}
data <- readRDS("data_gf.rds")
```

```
```{r}
abc <- perform_abc(data, tolerance = 0.05, method = "neuralnet")
```

```{r}
extract_summary(abc)
```

```{r}
hist(abc, "Ne_p1")
hist(abc, "Ne_p2")
hist(abc, "Ne_p3")
hist(abc, "Ne_p4")

plot(abc, param = "Ne_p1")
plot(abc, param = "Ne_p2")
plot(abc, param = "Ne_p3")
plot(abc, param = "Ne_p4")

plot(abc, param = "Tsplit_p1_p2")
plot(abc, param = "Tsplit_p2_p3")
plot(abc, param = "Tsplit_p3_p4")

plot(abc, param = "gf_p4_p2")
```

Plot posterior distributions of interest:

```{r}
plot_posterior(abc)
plot_posterior(abc, type = "Ne")
plot_posterior(abc, type = "Tsplit")
plot_posterior(abc, type = "gf", xlim = c(0, 1))
```


```{r}
truth_df <- data.frame(
  param = c("Ne_p1", "Ne_p2", "Ne_p3", "Ne_p4"),
  value = c(3000, 250, 5000, 1000)
)

plot_posterior(abc) +
  ggplot2::geom_vline(data = truth_df, ggplot2::aes(xintercept = value), linetype = 2)
```


```{r}
truth_df <- data.frame(
  param = c("Tsplit_p2_p3", "Tsplit_p1_p2"),
  value = c(6000, 2000)
)

plot_posterior(abc, type = "Tsplit") +
  ggplot2::geom_vline(data = truth_df, ggplot2::aes(xintercept = value), linetype = 2)
```


```{r}
inferred_model <- extract_model(abc, summary = "mode")

plot_model(inferred_model)
```