---
title: Custom SLiM or Python simulations
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom SLiM or Python simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.align = "center",
  fig.width = 8,
  fig.height = 6,
  dpi = 80
)

data_path <- here::here("inst/examples/custom_abc.rds")

devtools::load_all()
```

By default, _demografr_ uses the [_slendr_](https://github.com/bodkan/slendr) simulation framework for defining models and simulating data from them. But what if you need to do inference using your own scripts? Perhaps _slendr_'s opinionated interface doesn't allow you to do all that you need to do, such as using some of the powerful simulation features and options of "raw" [SLiM](https://messerlab.org/slim/) or [_msprime_](https://tskit.dev/msprime/docs/stable/intro.html)? This vignette explains how you can use any standard SLiM or _msprime_ script which produces a tree sequence file as its output as an engine in a standard _demografr_ inference pipeline.

First, let's load _demografr_ itself and also _slendr_ (which, in this vignette, will serve only for analyzing simulated tree-sequence data, but not simulations themselves).

```{r}
library(demografr)
library(slendr)

init_env(quiet = TRUE)
```

## Restrictions on user-defined simulation scripts

In order to be able to use a custom SLiM or msprime script with _demografr_, the script must conform to a couple of rules:

### 1. It must be runnable on the command-line as any other command-line script

- **For an msprime script**, this means something like `python <your script>.py <command-line arguments>`.

- **For a SLiM script**, this means something like `slim <command-line arguments> <your script>.slim`.

### 2. It must accept three mandatory arguments with these exact names

- `sequence_length`: the amount of sequence to simulated
- `recombination_rate`: the recombination rate along the simulated sequence (in units of crossovers per basepair pergeneration)
- `output_path`: the path where the simulated tree-sequence object will be saved

**For an msprime script**, these parameters can be specified via the Python built-in module `argparse`, and provided on the command-line as `--sequence_length <value>`, `--recombination_rate <value>`, and `--output_path <path>`. In the script itself, if you use the `argparse` module and have thus the values of provided arguments (for instance) in an `args` object, you can then refer to these arguments as `args.sequence_length`, etc.

**For a SLiM script**, these parameters can be specified via SLiM's standard way of supplying command-line arguments as `-d sequence_length=<value>`, `-d recombination_rate=<value>`, and `-d "output_path='<path>'"`. Importantly, note SLiM's format of specifying string arguments for the `output_path` argument. If you need further detail, see Section 20.2 of the SLiM manual. In the script itself, you can then refer to these arguments via constants `sequence_length`, `recombination_rate`, and `output_path`.

### 3. All model parameters must be provided as additional command-line arguments

You can refer to them in your script as you would to the mandatory arguments as described in the previous section.

**A useful check to see whether you script would make a valid engine for a _demografr_ analysis is to run it manually on the command line like this:**

```
python \
  <path to your Python script>                   \
  --sequence_length <number>                     \
  --recombination_rate <number>                  \
  --output_path <path to the output .trees file> \
  <... your model parameters ...>
```

Or, if you want to run a SLiM-powered ABC, like this:

```
slim \
  -d sequence_length=<number>                         \
  -d recombination_rate=<number>                      \
  -d "output_path='<path to the output .trees file>'" \
  <... your model parameters ...>
```

Then, if you get a `.trees` file, you're good to go!

**Note** that mutation rate is not among the list of mandatory parameters for your script. Neutral mutations are taken care of via tree-sequence mutation process after the simulation is finished. If this sounds weird to you, take a look at some tree-sequence-related materials like [this](https://tskit.dev/tutorials/no_mutations.html).

## ABC analysis using custom-defined simulation scripts

### Toy problem

Suppose that we intent to use ABC to infer the $N_e$ of a constant-sized population given that we observed the following value of nucleotide diversity:

```{r, echo=FALSE}
ts <- population("pop", N = 500, time = 10000) %>%
  compile_model(generation_time = 1, direction = "backward") %>%
  msprime(sequence_length = 50e6, recombination_rate = 1e-8, random_seed = 42) %>%
  ts_mutate(mutation_rate = 1e-8, random_seed = 42)
observed_diversity <- ts_diversity(ts, list(pop = ts_samples(ts)$name))
```

```{r}
observed_diversity
```

We want to infer the posterior distribution of $N_e$ but not via the normal _slendr_ interface (as shown [here](vignette-01-basics.html)) but using a simple SLiM or Python script.

**We acknowledge that this is a completely trivial example, not really worth spending so much effort on doing an ABC for.** The example was chosen because it runs fast and demonstrates all the features of _demografr_ that you would use regardless of the complexity of your model.

## Example pure SLiM and msprime script

To demonstrate the requirements 1-3 above in practice, let's perform ABC using each of those scripts as tree-sequence simulation engines.

Imagine we have the following two scripts which we want to use as simulation engines instead of _slendr_'s own functions `slim()` and `msprime()`. The scripts have only **one _model parameter_, $N_e$ of the single modelled population**. The **other parameters are mandatory**, just discussed in the previous section.

### SLiM script

```{r}
slim_script <- system.file("examples/custom.slim", package = "demografr")
```

```{r, comment="", echo=FALSE}
cat(readLines(slim_script), sep = "\n")
```

### Python script

```{r}
python_script <- system.file("examples/custom.py", package = "demografr")
```

```{R, comment="", echo=FALSE}
cat(readLines(python_script), sep = "\n")
```

**Note that the Python script specifies all model parameters (here, $N_e$) and mandatory arguments, via a command-line interface provided by the Python module `argparse`**.

**The SLiM script, on the other hand, uses SLiM's features for command-line specification of parameters and simply refers to each parameter by its symbolic name.** No other setup is necessary.

### Individual ABC components

Apart from the user-defined simulation SLiM and _msprime_ scripts, the components of our toy inference remains the same&mdash;we need to define the observed statistics, tree-sequence summary functions, and priors. We don't need a model function&mdash;that will be served by our custom script.

Here are the _demografr_ pipeline components (we won't be discussing them here because that's extensively taken care of [elsewhere](vignette-01-basics.html) in _demografr_'s documentation vignettes):

```{r}
# a single prior parameter
priors <- list(Ne ~ runif(10, 1000))

# a single observed statistic
observed <- list(diversity = observed_diversity)

# a single tree-sequence summary function producing output in the same
# format as the observed statistic table
functions <- list(
  diversity = function(ts) {
    ts_diversity(ts, list(pop = seq(0, 19))) # compute using 20 chromosomes
  }
)
```

(Note that because our simulated tree sequences won't be coming with _slendr_ metadata, we have to refer to individuals' chromosomes using numerical indices rather than _slendr_ symbolic names like we do in all of our other vignettes.)

### Simulating a testing tree sequence

As we explained elsewhere, a useful function for developing inference pipelines using _demografr_ is a function `simulate_ts()`, which normally accepts a _slendr_ model generating function and the model parameters (either given as priors or as a list of named values), and simulates a tree-sequence object. This function (as any other _demografr_ function operating with models) accepts our custom-defined simulation scripts in place of standard _slendr_ models.

For instance, we can simulate a couple of Mb of testing sequence from our SLiM script like this:

```{r}
ts_slim <- simulate_ts(slim_script, parameters = list(Ne = 1234),
                       sequence_length = 10e6, recombination_rate = 1e-8)
```

(Of course we could've plugged in the `python_script` instead.)

Now we can finally test our little toy tree-sequence summary function, verifying that we can indeed compute the summary statistic we want to:

```{r}
functions$diversity(ts_slim)
```

The function works as expected&mdash;we have a single population and want to compute nucleotide diversity in the whole population, and this is exactly what we get.

### Running an ABC analysis

Having all components of our pipeline set up we should, again, validate everything before we proceed to (potentially very costly) simulations. In the remainder of this vignette we'll only continue with the Python msprime custom script in order to save some computational time. That said, it's important to realize that you could use this sort of workflow for any kind of SLiM script, including very elaborate spatial simulations, non-WF models, and all kinds of phenotypic simulations too! **Although primarily designed to work with _slendr_, the _demografr_ package intends to fully support any kind of SLiM or _msprime_ simulation.**

Let's first validate all components of our pipeline:

```{r}
validate_abc(python_script, priors, functions, observed)
```

Looking good! Now let's first run ABC simulations. Again, note the use of the Python script where we would normally plug in a _slendr_ model function in place of the `model` argument:

```{r, echo=FALSE, eval=!file.exists(data_path)}
tstart <- Sys.time()
```

```{r, eval=!file.exists(data_path)}
library(future)
plan(multicore, workers = availableCores()) # parallelize across all 96 CPUs

data <- simulate_abc(
  model = python_script, priors, functions, observed, iterations = 1000,
  sequence_length = 1e6, recombination_rate = 1e-8, mutation_rate = 1e-8
)
```

```{r, echo=FALSE, eval=!file.exists(data_path)}
tend <- Sys.time()

tdelta <- as.numeric(difftime(tend, tstart, units = "secs"))
saveRDS(tdelta, here::here("inst/examples/custom_tdelta.rds"))
ncores <- future::availableCores()
saveRDS(ncores, here::here("inst/examples/custom_ncores.rds"))
```

```{r, echo=FALSE, eval=file.exists(data_path)}
tdelta <- readRDS(here::here("inst/examples/custom_tdelta.rds"))
ncores <- readRDS(here::here("inst/examples/custom_ncores.rds"))
```

```{r, echo=FALSE}
hours <- floor(tdelta / 3600)
minutes <- floor((tdelta - hours * 3600) / 60)
seconds <- round(tdelta - hours * 3600 - minutes * 60)
```

**The total runtime for the ABC simulations was `r paste(hours, "hours", minutes, "minutes", seconds, "seconds")` parallelized across `r ncores` CPUs.**

Once the simulations are finished, we can perform inference of the posterior distribution of the single parameter of our model, $N_e$:

```{r, eval=!file.exists(data_path)}
abc <- perform_abc(data, engine = "abc", tol = 0.03, method = "neuralnet")
```

```{r, echo=FALSE, eval=!file.exists(data_path)}
saveRDS(abc, data_path)
```

```{r, echo=FALSE, eval=file.exists(data_path)}
abc <- readRDS(data_path)
```

Having done so, we can again look at the summary statistics of the posterior distribution and also plot the results:

```{r}
extract_summary(abc)
```

```{r, custom_Ne_posterior}
plot_posterior(abc)
```

Again, in addition to _demografr_'s visualization functions (like `plot_posterior()` above) we have the full range of diagnostic plotting functions bundled with the _abc_ R package at our disposal:

```{r, custom_Ne_diagnostic}
plot(abc)
```